A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-09-02 main.tex
Notes to self

Different variance component estimation methods lead to two different forms of $\widehat{\bW}$ and LMM-Lasso variants: LMM-Lasso-Rakitsch and LMM-Lasso-ggmix. LMM-Lasso-Rakitsch uses a two-step procedure for estimating the variance components, $\sigmaee$ and $\delta := \sigmaee / \sigmagg$ \citep{Rakitsch2012}. These variance components and the corresponding weights are estimated once under the assumption of a null model. A Lasso model can then be fit on the transformed data to estimate $\boldsymbol{\beta}$ using standard software for coordinate descent, such as \texttt{glmnet} \citep{glmnet} or \texttt{ncvreg} \citep{ncvreg}. LMM-Lasso-ggmix iteratively updates $\widehat{\boldsymbol{\beta}}$ and the variance components via a block coordinate descent algorithm \citep{bhatnagar2019simultaneous}. This iterative update scheme necessitates the use of the corresponding \texttt{R} package \texttt{ggmix} \citep{ggmix}. We postpone further discussion of the differences between these methods to section \ref{sec:results}.

@@

This result also allows us to mathematically justify the statement from section \ref{sec:background}, that SNP-phenotype associations will be unbiased in the presence of environmental heterogeneity which does not correspond to population structure. We can quantify the proportion of the environmental effect which is in the column space of $\boldsymbol{P}_{\bX}$ using the ratio $||\bX \boldsymbol{\tau}||_2 / ||\bZ \bgamma||_2 \in [0, 1]$. In the case where $\bZ \bgamma$ is entirely independent of population structure, $\bZ \bgamma \in \mathcal{N}(\boldsymbol{P}_{\bX})$ and $\bX \boldsymbol{\tau} = \boldsymbol{0}$ so that $\mathbb{E}(\bbetaHat^*) = \bbeta$. Thus the SNP-phenotype associations will be unbiased, but the residual variance will be increased. Although we focus our attention on accurately estimating fixed SNP effects, we note that this result could lead to substantial bias in narrow-sense heritability estimates. In practical applications $\bZ \bgamma$ is unlikely to be entirely orthogonal to population structure. However, the degree to which it reflects population structure that can be inferred from $\bX$, as quantified by $||\bX \boldsymbol{\tau}||_2 / ||\bZ \bgamma||_2$, will impact the extent to which environmental heterogeneity confounds SNP-phenotype associations.

@@

Given the nature of environmental confounding in our setting of interest, and in order to assess these methodological performance in terms of estimation accuracy and bias specifically, we simulate according to a fixed $\bgamma$ framework using the following data-generating model
\begin{equation}
    \label{eqn:sim}
    \by = a \bX \bbeta + b \bZ \bgamma + c \beps,
\end{equation}
where $\by$ represents the pseudophenotype, and $a, b, c$ are scaling factors determined such that the variance of $\by$ is equal to 1 and is partitioned into genetic signal, environmental confounding, and random noise components,
\begin{equation}
    \label{eqn:parition_y}
    \mathbb{V}(\by) = \eta + (1 - \eta) \xi + (1-\eta)(1-\xi).
\end{equation}
$\eta$ defines the proportion of the variance of $\by$ attributable to additive genetic factors, commonly referred to as the narrow-sense heritability, $h^2 = \sigmagg/(\sigmagg + \sigmaee)$. $\eta$ also describes the data-generating model's signal to noise ratio (SNR). $\xi$ partitions the remaining variance of $\by$ into latent environmental heterogeneity and random noise. So, for example $\eta = 0.5$ and $\xi = 0.8$ corresponds to a 1:1 SNR, and means that 50\% of the variability in $\by$ is due to additive genetic effects, and 40\% due to environmental effects.

@@

\subsubsection{$\lambda$ selection}
The $\lambda$ that allowed for the first 50 variables to enter the model was selected for each method, where 50 was the true number of causal SNPs. This method for selecting $\lambda$ was chosen in an attempt to compare the performance of all methods fairly, and as a strategy to evaluate methods in terms of variable selection, as it allows for direct comparison of the first 50 variables each method lets into the model. Alternative $\lambda$ selection methods were also considered. Cross-validation has been found to be inadvisable for LMMs, and in the presence of non-inedpendent samples as is the case here. While information criterion methods such as BIC present another option for selecting $\lambda$, they may be unstable in high dimensions. In addition, because Lasso, PC-Lasso, and LMM-Lasso use different likelihoods an information-criterion-based approach would not lend itself to comparable results across methods.

@@

Finally, we considered two implementations of the LMM-Lasso method: LMM-Lasso-Rakitsch and LMM-Lasso-ggmix, which differ in their variance estimation procedures and computational implementations. Although we found that both methods perform similarly in terms of $\bbeta$ estimation accuracy, LMM-Lasso-ggmix has several drawbacks. Firstly, it suffers from estimation instability in $n \ge p$ settings. Secondly, although LMM-Lasso-Rakitsch and LMM-Lasso-ggmix perform comparably when $n << p$, LMM-Lasso-ggmix requires up to 60 times the computing time of LMM-Lasso-Rakitsch in these settings due to its iterative estimation procedure. The additional time required of LMM-Lasso-ggmix may be worthwhile if the objective of an analysis is to accurately estimate the narrow-sense heritability in $n << p$ settings. However, the LMM-Lasso-Rakitsch method is more robust to varying levels of environmental confounding and sample sizes and thus considered by the authors to be more broadly applicable.

>>>>

Different variance component estimation methods lead to different forms of $\widehat{\bW}$ and two distinct LMM-Lasso variants: LMM-Lasso-Rakitsch and LMM-Lasso-ggmix. LMM-Lasso-Rakitsch uses a two-step procedure for estimating the variance components, $\sigmaee$ and $\delta := \sigmaee / \sigmagg$ \citep{Rakitsch2012}. These variance components and the corresponding weights are estimated once under the assumption of a null model. A Lasso model can then be fit on the transformed data to estimate $\boldsymbol{\beta}$ using standard software for coordinate descent, such as \texttt{glmnet} \citep{glmnet} or \texttt{ncvreg} \citep{ncvreg}. LMM-Lasso-ggmix iteratively updates $\widehat{\boldsymbol{\beta}}$ and the variance components via a block coordinate descent algorithm \citep{bhatnagar2019simultaneous}. This iterative update scheme necessitates the use of the corresponding \texttt{R} package \texttt{ggmix} \citep{ggmix}. We postpone further discussion of the differences between these methods to section \ref{sec:results}.

@@

This result also allows us to mathematically justify the statement from section \ref{sec:background}, that SNP-phenotype associations will be unbiased in the presence of environmental heterogeneity which does not correspond to population structure. We can quantify the proportion of the environmental effect which is in the column space of $\boldsymbol{P}_{\bX}$ using the ratio $||\bX \boldsymbol{\tau}||_2 / ||\bZ \bgamma||_2 \in [0, 1]$. In the case where $\bZ \bgamma$ is entirely independent of population structure, $\bZ \bgamma \in \mathcal{N}(\boldsymbol{P}_{\bX})$ and $\bX \boldsymbol{\tau} = \boldsymbol{0}$ so that $\mathbb{E}(\bbetaHat^*) = \bbeta$. Thus the SNP-phenotype associations will be unbiased, but the residual variance will be increased. Although we focus our attention on accurately estimating fixed SNP effects, we note that this result could lead to substantial bias in narrow-sense heritability estimates. In practical applications $\bZ \bgamma$ is unlikely to be entirely orthogonal to population structure. However, the degree to which it reflects population structure that can be inferred from $\bX$, as quantified by $||\bX \boldsymbol{\tau}||_2 / ||\bZ \bgamma||_2$, will impact the extent to which environmental heterogeneity confounds SNP-phenotype associations. \anna{this calculation is still having issues for the admixture and independent supbopulation data - the ratio is slightly larger for the independent data - as we would expect - but only on the order of 1e-15... correlation of Xtau with Zgamma? euclidian distance between vectors?}

@@

Given the nature of environmental confounding in our setting of interest, and in order to assess these methodological performance in terms of estimation accuracy and bias specifically, we simulate according to a fixed $\bgamma$ framework using the following data-generating model
\begin{equation}
    \label{eqn:sim}
    \by = a \bX \bbeta + b \bZ \bgamma + c \beps,
\end{equation}
where $\by$ represents the pseudophenotype, and $a, b, c$ are scaling factors determined such that the variance of $\by$ is equal to 1 and is partitioned into genetic signal, environmental confounding, and random noise components,
\begin{equation}
    \label{eqn:parition_y}
    \mathbb{V}(\by) = \eta + (1 - \eta) \xi + (1-\eta)(1-\xi).
\end{equation}
$\eta$ defines the proportion of the variance of $\by$ attributable to additive genetic factors, commonly referred to as the narrow-sense heritability, $h^2 = \sigmagg/(\sigmagg + \sigmaee)$. $\eta$ also describes the data-generating model's signal to noise ratio (SNR). $\xi$ partitions the remaining variance of $\by$ into latent environmental heterogeneity and random noise. So, for example $\eta = 0.5$ and $\xi = 0.8$ corresponds to a 1:1 SNR, and means that 50\% of the variability in $\by$ is due to additive genetic effects, and 40\% due to unobserved environmental effects.

@@

\subsubsection{$\lambda$ selection}
The $\lambda$ that allowed for the first 50 variables to enter the model was selected for each method, where 50 was the true number of causal SNPs. This method for selecting $\lambda$ was chosen in an attempt to compare the performance of all methods fairly, and as a strategy to evaluate methods in terms of variable selection, as it allows for direct comparison of the first 50 variables each method lets into the model. Alternative $\lambda$ selection methods were also considered. Cross-validation has been found to be inadvisable for LMMs, and in the presence of non-inedpendent samples as is the case here. While information criterion methods such as BIC present another option for selecting $\lambda$, they may be unstable in high dimensions. In addition, because Lasso, PC-Lasso, and LMM-Lasso use different likelihoods, an information-criterion-based approach would not lend itself to comparable results across methods.

@@

Finally, we considered two implementations of the LMM-Lasso method: LMM-Lasso-Rakitsch and LMM-Lasso-ggmix, which differ in their variance estimation procedures and computational implementations. Although we found that both methods perform similarly in terms of $\bbeta$ estimation accuracy, LMM-Lasso-ggmix has several drawbacks. Firstly, it suffers from estimation instability in $n \ge p$ settings. Secondly, although LMM-Lasso-Rakitsch and LMM-Lasso-ggmix perform comparably when $n << p$, LMM-Lasso-ggmix requires up to 60 times the computing time of LMM-Lasso-Rakitsch in these settings due to its iterative estimation procedure. The additional time required of LMM-Lasso-ggmix may be worthwhile if the objective of an analysis is to accurately estimate the narrow-sense heritability in $n << p$ settings. However, the LMM-Lasso-Rakitsch method is more robust to varying levels of environmental confounding and sample sizes and thus considered by the authors to be more broadly applicable. \anna{talk about how ggmix seems to get stuck in a local minimum because it's esimating more parameters the optimization problem is not necessarily convex, this its prone to getting stuck. While this may make it fast, it makes it terrible for estimation accuracy. Reference iterations plots?}
